BarnebyLives: an R package to create herbarium specimen labels and clean spreadsheets

### 2025-03-01 manuscript/BarnebyLives An R package to create and qc herbarium labels.rmd
need troubleshoot label making

--- 
title: 'BarnebyLives: an R package to create herbarium specimen labels and clean spreadsheets'
author:  |
    |  $^1$Chicago Botanic Garden, 1000 Lake Cook Road, Glencoe, Illinois 60022, USA  
    |  $^2$Plant Biology and Conservation, Northwestern University, Evanston, Illinois 60208, USA  
    | Reed Clark Benkendorf$^1$^[Author for Correspondence: rbenkendorf@chicagobotanic.org]
abstract: |
  \noindent
  **Premise:** Depositing specimens to herbaria is a time consuming process. Many institutions have reduced the amount of funding for herbaria, and universities have reduced the amount of education dedicated towards collections and curatorial tasks. However, the continual generation of herbarium specimens are essential for current and future research in evolution and ecology. In order to faciliate the continued growth of herbaria BarnebyLives was developed as a tool to supplement collection notes, perform geographic and, taxonomic informatic processes, enact spell checks and other QC steps, produce labels, and submit digital data to increase the rate of accessioning specimens.  
  **Methods and Results:** BarnebyLives uses geospatial data from the U.S. Census Bureau to provide administrative jurisdiction information, and data from other sources, including the United States Geological Survey, to supplement collection notes by providing information on abiotic site conditions. It uses inhouse spell checks to verify the spelling of a collection at all taxonomic ranks, the IPNI standard author database to check standard author abbreviations, and the Royal Botanic Garden Kews 'Plants of the World Online' to check for nomenclatural innovations. Optionally the package writes driving directions to sites using Google Maps. The package outputs data in a tabular format, as well as a spatial format, for review by the user before rendering labels using LaTeX.       
  **Conclusions:** BarnebyLives provides accurate political and physical information, reduces typos, provides users the most current taxonomic opinions, generates driving directions to sites, and produces aesthetically appealing labels and shipping manifests in a matter of minutes.     
keywords: |
  herbarium, natural history museum, collections, geospatial, R, LaTeX   
output:
  word_document: default
  toc: no
  pdf_document: default
csl: "../citations/american-journal-of-botany.csl"
bibliography: ../citations/citations.bib
link-citations: yes
fig_caption: yes
always_allow_html: yes
header-includes:
- \usepackage{endfloat}
- \usepackage{setspace}\doublespacing
- \usepackage{lineno}
- \linenumbers
- \usepackage[width=\textwidth]{caption}
- \usepackage{wrapfig}
- \usepackage[export]{adjustbox}
- \usepackage{rotating}
---

@@

We anticipate that for the typical user BarnebyLives will require less than a couple gigabytes of memory (ours covering all of the conterminous Western U.S. at 90m resolution is ~16 GiB), while the processing takes relatively little RAM, hence we believe installations can work on hardware as limited as Chromebooks, while having the data stored entirely on thumb-drives.
The final steps of BarnebyLives, generating the labels, requires working installations of R Markdown, a LaTeX installation (e.g. [pdfTeX](https://www.tug.org/applications/pdftex/), [LuaTeX](https://www.luatex.org/), [XeLaTeX](https://www.overleaf.com/learn/latex/XeLaTeX)), and the open source command line tools [pdfjam](https://github.com/rrthomas/pdfjam) and [pdftk](https://linux.die.net/man/1/pdftk).
While these steps are run through a shell scripting language like bash, we have wrapped them in R functions which bypass the need to enter the commands directly into a terminal. 
Unfortunately, we have not found Windows alternatives to pdfam and pdftek so we are unable to offer the final label generating functionality on that operating system. 
BarnebyLives was iteratively developed based on data submitted by roughly 20 seasonal field botanists. 
Essentially, additions to the code have been continually made as the developers were exposed to more idiosyncrasies of collection notes and data entry.

## Usage
All steps of BarnebyLives except for label generation are run within the freely available Rstudio. 
Data may be read in from any common spreadsheet management system or database connection such as Excel, or free of charge alternatives such as: LibreOffice, OpenOffice, or via the cloud on Googlesheets. 
The latter two options are documented here and in package vignettes, detailed descriptions of the required and suggested input columns are located on the Github page (https://github.com/sagesteppe/BarnebyLives *'Input Data Column Names'*) and over *3*00 real-world examples are on a Google Sheets accessible from the page.
BarnebyLives is atypical of R packages in that it requires a considerable amount of data to operate (Table 1).
Virtually all the on-disk memory associated with these data are used in storing spatial data, setting up a local instance of the program - at whichever scale a user desires (see Figure XX) is fully documented in the package documentation. 
Functions which require the on-disk data require a path to the data as an argument. 
Manually supplying the path argument allows for users to judiciously decide a storage location suitable for their needs.

We anticipate BarnebyLives usage will requite less than a couple gigabytes of memory (ours covering all of the conterminous Western U.S. at 90m resolution is ~16 GiB), and the processing takes relatively little RAM, hence we believe installations can work on hardware as limited as Chromebooks, while having the data stored entirely on thumb-drives.
The final steps of BarnebyLives, generating the labels require working installations of Rmarkdown, a LaTeX installation (e.g. [pdfTeX](https://www.tug.org/applications/pdftex/), [LuaTeX](https://www.luatex.org/), [XeLaTeX](https://www.overleaf.com/learn/latex/XeLaTeX)), and the open source command line tools [pdfjam](https://github.com/rrthomas/pdfjam) and [pdftk](https://linux.die.net/man/1/pdftk).
While these steps are run through a shell scripting language like bash, we have wrapped them in R functions which bypass the need to enter the commands directly into a terminal.
Several commands in BarnebyLives require the output from previous functions, and a workflow which satisfies these requirements is presented in Figure 1.

@@

time_associate_dropper <- system.time({ # remove the focal taxon from the noted associates
  data <- associate_dropper(data, 'Full_name')
})

@@

saveRDS(time_directions_grabber, file = 'data/processed/time_directions_grabber')
saveRDS(data, file = 'data/processed/data_w_Google_Maps')
```

@@

>>>>

--- 
title: 'BarnebyLives: an R package to create herbarium specimen labels and clean spreadsheets'
author:  |
    |  $^1$Chicago Botanic Garden, 1000 Lake Cook Road, Glencoe, Illinois 60022, USA  
    |  $^2$Plant Biology and Conservation, Northwestern University, Evanston, Illinois 60208, USA  
    | Reed Clark Benkendorf$^1$$^,$$^2$^[Author for Correspondence: rbenkendorf@chicagobotanic.org], Jeremie B. Fant$^1$$^,$$^2$
abstract: |
  \noindent
  **Premise:** Depositing specimens to herbaria is a time consuming process. Many institutions have reduced the amount of funding for herbaria, and universities have reduced the amount of education dedicated towards collections and curatorial tasks. However, the continual generation of herbarium specimens are essential for current and future research in evolution and ecology. In order to faciliate the continued growth of herbaria BarnebyLives was developed as a tool to supplement collection notes, perform geographic and, taxonomic informatic processes, enact spell checks and other QC steps, produce labels, and submit digital data to increase the rate of accessioning specimens.  
  **Methods and Results:** BarnebyLives uses geospatial data from the U.S. Census Bureau to provide administrative jurisdiction information, and data from other sources, including the United States Geological Survey, to supplement collection notes by providing information on abiotic site conditions. It uses inhouse spell checks to verify the spelling of a collection at all taxonomic ranks, the IPNI standard author database to check standard author abbreviations, and the Royal Botanic Garden Kews 'Plants of the World Online' to check for nomenclatural innovations. Optionally the package writes driving directions to sites using Google Maps. The package outputs data in a tabular format, as well as a spatial format, for review by the user before rendering labels using LaTeX.       
  **Conclusions:** BarnebyLives provides accurate political and physical information, reduces typos, provides users the most current taxonomic opinions, generates driving directions to sites, and produces aesthetically appealing labels and shipping manifests in a matter of minutes.     
keywords: |
  herbarium, natural history museum, collections, geospatial, R, LaTeX   
output:
  pdf_document: default
  toc: no
  word_document: default
csl: "../citations/american-journal-of-botany.csl"
bibliography: ../citations/citations.bib
link-citations: yes
fig_caption: yes
always_allow_html: yes
header-includes:
- \usepackage{endfloat}
- \usepackage{setspace}\doublespacing
- \usepackage{lineno}
- \linenumbers
- \usepackage[width=\textwidth]{caption}
- \usepackage{wrapfig}
- \usepackage[export]{adjustbox}
- \usepackage{rotating}
---

@@

We anticipate that for the typical user BarnebyLives will require less than a couple gigabytes of memory (ours covering all of the conterminous Western U.S. at 90m resolution is ~16 GiB), while the processing takes relatively little RAM, hence we believe installations can work on hardware as limited as Chromebooks, while having the data stored entirely on thumb-drives.
The final steps of BarnebyLives, generating the labels, requires working installations of R Markdown, a LaTeX installation (e.g. [pdfTeX](https://www.tug.org/applications/pdftex/), [LuaTeX](https://www.luatex.org/), [XeLaTeX](https://www.overleaf.com/learn/latex/XeLaTeX)), and the open source command line tools [pdfjam](https://github.com/rrthomas/pdfjam) and [pdftk](https://linux.die.net/man/1/pdftk).
While these steps are run through a shell scripting language like bash, we have wrapped them in R functions which bypass the need to enter the commands directly into a terminal. 
Unfortunately, we have not found Windows alternatives to pdfam and pdftek so we are unable to offer the final label generating functionality on that operating system. 
BarnebyLives was iteratively developed based on data submitted by roughly 20 seasonal field botanists. 
Essentially, additions to the code have been continually made as the developers were exposed to more idiosyncrasies of collection notes and data entry. 
Several commands in BarnebyLives require the output from previous functions, and a workflow which satisfies these requirements is presented in Figure 1.

@@

time_associate_dropper <- system.time({ # remove the focal taxon from the noted associates
  data <- associate_dropper(data, 'Full_name', col  = 'Associates')
})

time_associate_dropper <- system.time({ # remove the focal taxon from the noted associates
  data <- associate_dropper(data, 'Full_name', col  = 'Vegetation')
})

@@

saveRDS(time_directions_grabber, file = 'data/processed/time_directions_grabber')
saveRDS(data, file = 'data/processed/data_w_Google_Maps')

```{r accept or reject changes, eval = F}

# for this example we will just ACCEPT all changes blindly. 
# basically we just want some data for running timing exercises on the label making.

# just overwrite the original Associates and Vegetation columns
data <- data %>% 
  select(-Associates, -Vegetation) |>
  rename(Associates = 'SpellCk.Associates', Vegetation = 'SpellCk.Vegetation')

# honestly just overwrite these too. 
data <- data %>% 
  select(-SpellCk.gGRP, -Match, -SpellCk.taxon_rank, 
        -Infrarank, -Infraspecies) |>
  rename(Full_name = SpellCk.taxon_name, Genus = SpellCk.genus, Epithet = SpellCk.species, 
         Infrarank = SpellCk.infraspecific_rank, Infraspecies = SpellCk.infraspecies) %>% 
  mutate(
    Slope = slope, 
    Aspect = aspect
  ) |>
  unite(Name_authority, 
        Genus, Epithet, Binomial_authority, Infrarank, Infraspecies, Infraspecific_authority,
        remove = F, na.rm = T, sep = ' ') |>
  sf::st_drop_geometry()

# also save a spreadsheet of the cleaned data, we'll use this to create the labels. 
write.csv(data, file = 'data/processed/cleaned_data.csv', row.names = FALSE)
```

@@

```{r Create Labels}
processed <- read.csv('data/processed/cleaned_data.csv') %>% 
  mutate(UNIQUEID = paste0(Primary_Collector, Collection_number), 
         Coordinate_Uncertainty = '+/- 5m') %>% 
  data.frame()

processed <- data.frame( apply(processed, 2, as.character) )
processed <- mutate(processed, Collection_number = as.numeric(Collection_number))

dir.create('./labels/raw')
p <- '/media/steppe/hdd/Barneby_Lives-dev/manuscript/labels/raw'

purrr::walk(
  .x = processed$Collection_number,
  ~ rmarkdown::render(
    input = './labels/skeleton.Rmd',
    output_file = file.path(p, glue::glue("{.x}.pdf")),
    params = list(Collection_number = {.x})
  )
)
```

### 2025-03-01 manuscript/labels/skeleton.Rmd
need troubleshoot label making

---
output: pdf_document
params:
  Collection_number: Collection_number
  UNIQUEID: UNIQUEID
geometry: paperheight=4in,paperwidth=4in,margin=0.1in
header-includes:
  - \pagenumbering{gobble}
  - \usepackage{setspace}
---

@@

```{r load data}
data <- readRDS('../data/processed/data_w_Google_Maps') |>
  sf::st_drop_geometry() |>
  dplyr::filter(Collection_number == params$Collection_number) |>  
  sf::st_drop_geometry()
```

\begin{center}
\textbf{\LARGE Flora of Western North America}
\end{center}
\begin{flushright}
\uppercase{`r data$Family`}
\end{flushright}
\begingroup
\large
`r writer(paste(data$Genus, data$Epithet), italics = TRUE)` `r data$Binomial_authority` `r writer(data$Infraspecific_rank)` `r writer(data$Infraspecies, italics = TRUE)` `r writer(data$Infraspecific_authority)`  
\normalsize
\endgroup

\vspace*{\fill}
\begingroup
`r data$latitude_dd` `r data$longitude_dd` (`r data$Datum`). `r data$Gen`. `r writer(data$Site)`

`r writer(data$Habitat)` `r data$physical_environ`  
Veg.: `r species_font(data$Vegetation)` `r associates_writer(data$Associates)`
`r writer(data$Notes)`  
\endgroup
\begingroup
\begin{spacing}{0.1}
{\scriptsize `r data$Directions_BL`}
\end{spacing}
\endgroup

@@

\begin{center}
{\footnotesize Collected under the auspices of the Bureau of Land Management}
\end{center}

>>>>

---
output: pdf_document
params:
  Collection_number: Collection_number
geometry: paperheight=4in,paperwidth=4in,margin=0.1in
header-includes:
  - \pagenumbering{gobble}
  - \usepackage{setspace}
---

@@

```{r load data}
data <- read.csv('../data/test_data.csv') |>
  dplyr::mutate(
    Collection_number = as.numeric(Collection_number), 
    Coordinate_Uncertainty = '5m') |>
  dplyr::filter(Collection_number == params$Collection_number)
```

\begin{center}
\textbf{\LARGE}
\end{center}
\begin{flushright}
\uppercase{`r data$Family`}
\end{flushright}
\begingroup
\large
`r writer(paste(data$Genus, data$Epithet), italics = TRUE)` `r writer(data$Binomial_authority)` `r writer(data$Infrarank)` `r writer(data$Infraspecies, italics = TRUE)` `r writer(data$Infraspecific_authority)`  
\normalsize
\endgroup

\vspace*{\fill}
\begingroup
`r data$Gen`. `r writer(data$Site)` `r data$latitude_dd` `r data$longitude_dd` (`r data$Datum` `r writer(data$Coordinate_Uncertainty)`).

`r writer(data$Habitat)`. `r data$physical_environ`  
Veg.: `r species_font(data$Vegetation)` `r associates_writer(data$Associates)`
`r writer(data$Notes)`  
\endgroup
\begingroup
\begin{spacing}{0.1}
{\scriptsize `r data$Directions_BL`}
\end{spacing}
\endgroup

@@
